---
title: "Analysis"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
library(kyotil)
library(lme4)

# set drugs variable
dat=read.csv("fig5b.csv", row.names=1)
drugs=rownames(dat)
names(drugs)=drugs
```

## Intro

The IC50 is absolute IC50 not relative IC50. IC50 are log transformed in regression models and reported as ratios (not log ratio) in comparison. 

The unit of max inhibition is percentage and rounded to integer percentages as effect sizes for half a percentage point is not considered scientifically relevant. The maximum inhibition is not log transformed in analysis.

Ali and submerged are from the same donor.

NGC is excluded from the study drugs.

<!-- Coding note: log10 should not be used in the analysis. The data is log transformed and the analysis is based on exp transformed effects. Geometric means are computed by log transform followed by mean followed by exp transformation. -->

<!-- optional: improve axis labels -->


## Fig 2

The Wilcoxon two-sample test results are similar to t-test results.

```{r, echo=F, message=F}
dat=read.csv("fig2.csv")

pvals=sapply (1:4, function (i) {
  tmp=dat[,i*2+(-1:0)]
  wilcox.test(tmp[,1], tmp[,2])$p.value
})
pvals
```


## Fig 4/S7

Using linear mixed models to compare cell types to account for dependent structure of multiple donors. 


```{r, echo=F, message=F}

# multiple time points

for (i in 1:4) {
    
  if (i==1) {
    dat=read.csv("fig4.csv", row.names=1)
    print("Fig 4")
  } else if(i==2) {
    dat=read.csv("figS7_20.csv", row.names=1)
    print("Fig S7 20h")
  } else if (i==3) {
    dat=read.csv("figS7_36.csv", row.names=1)
    print("Fig S7 36h")
  } else {
    dat=read.csv("figS7_48.csv", row.names=1)
    print("Fig S7 48h")
  }
  
  dat.1=cbind(trt="Vero", t(dat[1,1:5]))
  colnames(dat.1)[2]="y"
  
  # remove the first row, the vero data
  dat = dat[-1,]
  
  # turn the data into a long format
  dat.2 = myreshapelong(dat[,c("k"%.%(1:5))], cols.to.be.stacked="k"%.%(1:5), label.cols.to.be.stacked="replicate", 
                        new.col.name="y")
  
  dat.3 = myreshapelong(dat[,c("f"%.%(1:3))], cols.to.be.stacked="f"%.%(1:3), label.cols.to.be.stacked="replicate", 
                        new.col.name="y")
  
  newdata = rbind(cbind(dat.2, Trt="K"), cbind(dat.3, Trt="F"), cbind(replicate="v"%.%1:5, y=dat.1[,"y"], id="0", Trt="V"))
  newdata$y=as.numeric(newdata$y)
  
  
  # fit a linear mixed model to compare two treatment options: Keratinocytes and V
  fit = lmer(log(y) ~ Trt + (1|id), data=subset(newdata, Trt %in% c("K", "V")))
  #summary(fit)
  
  # fit a linear mixed model to compare two treatment options: F and V
  fit2 = lmer(log(y) ~ Trt + (1|id), data=subset(newdata, Trt %in% c("F", "V")))
  #summary(fit2)
  
  # fit a linear mixed model to compare two treatment options: F and K
  # note that y is already log transformed in the data
  fit3 = lmer(log(y) ~ Trt + (1|id), data=subset(newdata, Trt %in% c("F", "K")))
  #summary(fit2)
  
  tab=getFormattedSummary(list("Keratinocytes-Vero"=fit, "Fibroblasts-Vero"=fit2, "Fibroblasts-Keratinocytes"=fit3), robust=T, p.digits=4, exp=T)
  print(tab[-1,])
  
  #tab=getFormattedSummary(list("Keratinocytes-V"=fit, "Fibroblasts-V"=fit2), robust=T, p.digits=4, type=10)
  #tab[-1,]

}

```



\newpage
## Fig 5b Fibroblasts-Keratinocytes comparison - IC50

Each point is a single technical replicate.

```{r, echo=F, fig.width=10, fig.height=3.1}
dat=read.csv("fig5b.csv", row.names=1)

lim=range(dat, na.rm=T)
cex=1.5

par(mfrow=c(1,3))
par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

plot(dat[,1], dat[,1+9], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Fibroblasts", main="IC50, Donor 3")
points(dat[,2], dat[,2+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
points(dat[,3], dat[,3+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
abline(0,1)
mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))

plot(dat[,1+1], dat[,1+9+1], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Fibroblasts", main="IC50, Donor 4")
points(dat[,2+1], dat[,2+9+1], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
points(dat[,3+1], dat[,3+9+1], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
abline(0,1)
mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))

plot(dat[,1+2], dat[,1+9+2], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Fibroblasts", main="IC50, Donor 5")
points(dat[,2+2], dat[,2+9+2], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
points(dat[,3+2], dat[,3+9+2], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
abline(0,1)
mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))
```

We fit a linear mixed model with a donor random effect for each drug to compare the IC50 of keratinocytes and fibroblasts. Multitesting-adjusted P values are shown below.

```{r, echo=F, message=F}
dat=read.csv("fig5b.csv", row.names=1)

fits=lapply (drugs, function (i) {
  tmp=data.frame(rbind(cbind("K", unlist(dat[i,1:9])), cbind("F", unlist(dat[i,1:9+9,drop=T]))))
  tmp$donor=rep(3:5, each=3)
  names(tmp)[1]="Trt"
  names(tmp)[2]="y"
  tmp[[2]]=as.numeric(tmp[[2]])
  
  capture.output({fit = lmer(log(y) ~ Trt + (1|donor), data=tmp)}, type="message")
  fit
})


tab=getFormattedSummary(fits, robust=T, p.digits=4, type=11, p.adj.method="holm", rows=2)
tab[order(as.numeric(tab))]
```


\newpage
### Fig S10. Donor effects

Each point is the average of three technical replicates.

```{r, echo=F, fig.width=8.5, fig.height=8}
dat=read.csv("fig5b.csv", row.names=1)

# take means across technical replicates
dat.1 = sapply (1:6, function (i) exp(apply(log(dat[,1:3+(i-1)*3]), 1, mean, na.rm=T))) 


lim=range(dat.1, na.rm=T)
cex=1.5

par(mfrow=c(2,2))
par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

plot(dat.1[,1], dat.1[,2], log="xy", pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 3", ylab="Keratinocytes, Donor 4", main="IC50")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))

plot(dat.1[,1], dat.1[,3], log="xy", pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 3", ylab="Keratinocytes, Donor 5", main="IC50")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))

plot(dat.1[,4], dat.1[,5], log="xy", pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 3", ylab="Fibroblasts, Donor 4", main="IC50")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))

plot(dat.1[,4], dat.1[,6], log="xy", pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 3", ylab="Fibroblasts, Donor 5", main="IC50")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))


# lim=range(dat, na.rm=T)
# cex=1.5
# 
# par(mfrow=c(2,2))
# par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))
# 
# plot(dat[,1], dat[,4], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 3", ylab="Keratinocytes, Donor 4", main="IC50")
# points(dat[,2], dat[,5], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# points(dat[,3], dat[,6], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# abline(0,1)
# mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))
# 
# plot(dat[,1], dat[,4+3], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 3", ylab="Keratinocytes, Donor 5", main="IC50")
# points(dat[,2], dat[,5+3], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# points(dat[,3], dat[,6+3], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# abline(0,1)
# mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))
# 
# plot(dat[,1+9], dat[,4+9], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 3", ylab="Fibroblasts, Donor 4", main="IC50")
# points(dat[,2+9], dat[,5+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# points(dat[,3+9], dat[,6+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# abline(0,1)
# mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))
# 
# plot(dat[,1+9], dat[,4+3+9], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 3", ylab="Fibroblasts, Donor 5", main="IC50")
# points(dat[,2+9], dat[,5+3+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# points(dat[,3+9], dat[,6+3+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# abline(0,1)
# mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))
```


Fit linear mixed models with donors as fixed effects and drug as a random effect and test the null hypothesis that the donor effects, 3 vs 4 and 3 vs 5, are both 0. P values are shown below.

```{r, echo=F, message=F}
dat=read.csv("fig5b.csv", row.names=1)

for(j in 0:1) {
  if (j==0) print("Keratinocytes") else print("Fibroblasts")
  tmp=data.frame(rbind(cbind(rownames(dat), 3, unlist(dat[,1+9*j])),
                       cbind(rownames(dat), 3, unlist(dat[,2+9*j])),
                       cbind(rownames(dat), 3, unlist(dat[,3+9*j])),
                       cbind(rownames(dat), 4, unlist(dat[,4+9*j])),
                       cbind(rownames(dat), 4, unlist(dat[,5+9*j])),
                       cbind(rownames(dat), 4, unlist(dat[,6+9*j])),
                       cbind(rownames(dat), 5, unlist(dat[,7+9*j])),
                       cbind(rownames(dat), 5, unlist(dat[,8+9*j])),
                       cbind(rownames(dat), 5, unlist(dat[,9+9*j]))
                       ))
  
  names(tmp)=c("drug","donor","y")
  tmp[[3]]=as.numeric(tmp[[3]])
  
  fit = lmer(log(y) ~ donor + (1|drug), data=tmp)
  est=unlist(coef(fit)[[1]][1,]) # all rows are identical in the second and third columns
  
  cat("Donor 3 vs 4 p value: ", pchisq((est[2] %*% solve(vcov(fit)[2,2]) %*% est[2])[1,1], 1, lower=F), "\n")
  cat("Donor 3 vs 5 p value: ", pchisq((est[3] %*% solve(vcov(fit)[3,3]) %*% est[3])[1,1], 1, lower=F), "\n")
  cat("Donor 4 vs 5 p value: ", pchisq(((est[2]-est[3]) %*% solve(c(1,-1)%*%vcov(fit)[2:3,2:3]%*%c(1,-1)) %*% (est[2]-est[3]))[1,1], 1, lower=F), "\n")
  
  cat("Overall p value: ", pchisq((est[-1] %*% solve(vcov(fit)[2:3,2:3]) %*% est[-1])[1,1], 2, lower=F), "\n")
}
```


This time, including comparion of Donor 4 and 5

```{r, echo=F, fig.width=8.5, fig.height=8}
dat=read.csv("fig5b.csv", row.names=1)

# take means across technical replicates
dat.1 = sapply (1:6, function (i) exp(apply(log(dat[,1:3+(i-1)*3]), 1, mean, na.rm=T))) 


lim=range(dat.1, na.rm=T)
cex=1.5

par(mfrow=c(2,2))
par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

plot(dat.1[,2], dat.1[,3], log="xy", pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 4", ylab="Keratinocytes, Donor 5", main="IC50")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))

empty.plot()


plot(dat.1[,5], dat.1[,6], log="xy", pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 4", ylab="Fibroblasts, Donor 5", main="IC50")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))


```

\newpage
## Fig 5c Fibroblasts-Keratinocytes comparison - Max Inhibition

Each point is a single technical replicate.

```{r, echo=F, fig.width=10, fig.height=3.1}
dat=read.csv("fig5c.csv", row.names=1)

lim=range(dat, na.rm=T)
cex=1.5

par(mfrow=c(1,3))
par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

plot(dat[,1], dat[,1+9], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Fibroblasts", main="Max Inhibition, Donor 3")
points(dat[,2], dat[,2+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
points(dat[,3], dat[,3+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
abline(0,1)
mylegend(rownames(dat), x=4, col=1:nrow(dat), pch=1:nrow(dat))

plot(dat[,1+1], dat[,1+9+1], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Fibroblasts", main="Max Inhibition, Donor 4")
points(dat[,2+1], dat[,2+9+1], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
points(dat[,3+1], dat[,3+9+1], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
abline(0,1)
mylegend(rownames(dat), x=4, col=1:nrow(dat), pch=1:nrow(dat))

plot(dat[,1+2], dat[,1+9+2], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Fibroblasts", main="Max Inhibition, Donor 5")
points(dat[,2+2], dat[,2+9+2], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
points(dat[,3+2], dat[,3+9+2], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
abline(0,1)
mylegend(rownames(dat), x=4, col=1:nrow(dat), pch=1:nrow(dat))
```

Multitesting-adjusted P values are shown below.

```{r, echo=F, message=F}
dat=read.csv("fig5c.csv", row.names=1)
dat=round(dat, 0)

fits=lapply (drugs, function (i) {
  tmp=data.frame(rbind(cbind("K", unlist(dat[i,1:9])), cbind("F", unlist(dat[i,1:9+9,drop=T]))))
  tmp$donor=rep(3:5, each=3)
  names(tmp)[1]="Trt"
  names(tmp)[2]="y"
  tmp[[2]]=as.numeric(tmp[[2]])
  
  capture.output({fit = lmer(y ~ Trt + (1|donor), data=tmp)}, type="message")
  fit
})

tab=getFormattedSummary(fits, robust=T, p.digits=4, type=11, p.adj.method="holm", rows=2)
tab[order(as.numeric(tab))]
```



\newpage
### Fig S10. Donor effects

Each point is the average of three technical replicates.

```{r, echo=F, fig.width=8.5, fig.height=8}
dat=read.csv("fig5c.csv", row.names=1)
dat=round(dat, 0)
# take means across technical replicates
dat.1 = sapply (1:6, function (i) (apply((dat[,1:3+(i-1)*3]), 1, mean, na.rm=T))) 

lim=range(dat.1, na.rm=T)
cex=1.5

par(mfrow=c(2,2))
par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

plot(dat.1[,1], dat.1[,2], pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 3", ylab="Keratinocytes, Donor 4", main="Max Inhibition")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))

plot(dat.1[,1], dat.1[,3], pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 3", ylab="Keratinocytes, Donor 5", main="Max Inhibition")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))

plot(dat.1[,4], dat.1[,5], pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 3", ylab="Fibroblasts, Donor 4", main="Max Inhibition")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))

plot(dat.1[,4], dat.1[,6], pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 3", ylab="Fibroblasts, Donor 5", main="Max Inhibition")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))


# lim=range(dat, na.rm=T)
# cex=1.5
# 
# par(mfrow=c(2,2))
# par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))
# 
# plot(dat[,1], dat[,4], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 3", ylab="Keratinocytes, Donor 4", main="Max Inhibition")
# points(dat[,2], dat[,5], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# points(dat[,3], dat[,6], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# abline(0,1)
# mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))
# 
# plot(dat[,1], dat[,4+3], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 3", ylab="Keratinocytes, Donor 5", main="Max Inhibition")
# points(dat[,2], dat[,5+3], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# points(dat[,3], dat[,6+3], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# abline(0,1)
# mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))
# 
# plot(dat[,1+9], dat[,4+9], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 3", ylab="Fibroblasts, Donor 4", main="Max Inhibition")
# points(dat[,2+9], dat[,5+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# points(dat[,3+9], dat[,6+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# abline(0,1)
# mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))
# 
# plot(dat[,1+9], dat[,4+3+9], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 3", ylab="Fibroblasts, Donor 5", main="Max Inhibition")
# points(dat[,2+9], dat[,5+3+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# points(dat[,3+9], dat[,6+3+9], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# abline(0,1)
# mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))
```



Fit linear mixed models with donors as fixed effects and drug as a random effect and test the null hypothesis that the donor effects, 3 vs 4 and 3 vs 5, are both 0. P values are shown below.

```{r, echo=F, message=F}
dat=read.csv("fig5c.csv", row.names=1)
dat=round(dat, 0)

for(j in 0:1) {
  if (j==0) print("Keratinocytes") else print("Fibroblasts")
  tmp=data.frame(rbind(cbind(rownames(dat), 3, unlist(dat[,1+9*j])),
                       cbind(rownames(dat), 3, unlist(dat[,2+9*j])),
                       cbind(rownames(dat), 3, unlist(dat[,3+9*j])),
                       cbind(rownames(dat), 4, unlist(dat[,4+9*j])),
                       cbind(rownames(dat), 4, unlist(dat[,5+9*j])),
                       cbind(rownames(dat), 4, unlist(dat[,6+9*j])),
                       cbind(rownames(dat), 5, unlist(dat[,7+9*j])),
                       cbind(rownames(dat), 5, unlist(dat[,8+9*j])),
                       cbind(rownames(dat), 5, unlist(dat[,9+9*j]))
                       ))
  
  names(tmp)=c("drug","donor","y")
  tmp[[3]]=as.numeric(tmp[[3]])
  
  # myboxplot(y~drug, data=tmp, pch=tmp$donor)
  
  fit = lmer(y ~ donor + (1|drug), data=tmp)
  est=unlist(coef(fit)[[1]][1,])
  
  cat("Donor 3 vs 4 p value: ", pchisq((est[2] %*% solve(vcov(fit)[2,2]) %*% est[2])[1,1], 1, lower=F), "\n")
  cat("Donor 3 vs 5 p value: ", pchisq((est[3] %*% solve(vcov(fit)[3,3]) %*% est[3])[1,1], 1, lower=F), "\n")
  cat("Donor 4 vs 5 p value: ", pchisq(((est[2]-est[3]) %*% solve(c(1,-1)%*%vcov(fit)[2:3,2:3]%*%c(1,-1)) %*% (est[2]-est[3]))[1,1], 1, lower=F), "\n")
  
  cat("Overall p value: ", pchisq((est[-1] %*% solve(vcov(fit)[2:3,2:3]) %*% est[-1])[1,1], 2, lower=F), "\n")
  
}
```


This time, including comparion of Donor 4 and 5

```{r, echo=F, fig.width=8.5, fig.height=8}
dat=read.csv("fig5c.csv", row.names=1)
dat=round(dat, 0)
# take means across technical replicates
dat.1 = sapply (1:6, function (i) (apply((dat[,1:3+(i-1)*3]), 1, mean, na.rm=T))) 

lim=range(dat.1, na.rm=T)
cex=1.5

par(mfrow=c(2,2))
par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

plot(dat.1[,2], dat.1[,3], pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes, Donor 4", ylab="Keratinocytes, Donor 5", main="Max Inhibition")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))

empty.plot()

plot(dat.1[,5], dat.1[,6], pch=1:nrow(dat.1), col=1:nrow(dat.1), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts, Donor 4", ylab="Fibroblasts, Donor 5", main="Max Inhibition")
abline(0,1)
mylegend(rownames(dat.1), x=1, col=1:nrow(dat.1), pch=1:nrow(dat.1))

empty.plot()
```

\newpage
## Fig 6a Four tissue models comparison - IC50

Each point in the figure below represents the average of technical replicates for each donor. Thus, for Ali and Submerged, there is only one unique value for each drug.

```{r, echo=F, fig.width=9.6, fig.height=6}
dat=read.csv("fig5b.csv", row.names=1)
# take means across technical replicates
# dat.1 = sapply (1:6, function (i) exp(apply(log(dat[,1:3+(i-1)*3]), 1, mean, na.rm=T))) 
# take means across technical and donor replicates
dat.1 = sapply (1:2, function (i) exp(apply(log(dat[,1:9+(i-1)*9]), 1, mean, na.rm=T)))

# add ali and submerged data
dat=read.csv("fig6a.csv", row.names=1)
# take means across technical replicates
submerged.mean=exp(apply(log(dat[,1:8]), 1, mean, na.rm=T))
ali.mean=exp(apply(log(dat[,9:13]), 1, mean, na.rm=T))

dat=cbind(dat.1, 
          submerged.mean, 
          ali.mean)


colnames(dat)=c("K", 
                "F",
                "S",
                "A"
                )
mywrite.csv(dat, "fig6aplot", row.names = T)

lim=range(dat, na.rm=T)
cex=1.5

par(mfcol=c(2,3))
par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

# Fibroblasts-Keratinocytes comparison
plot(dat[,"K"], dat[,"F"], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Fibroblasts", main="IC50, Fibroblasts-Keratinocytes")
# points(dat[,2], dat[,5], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
# points(dat[,3], dat[,6], col=1:nrow(dat), pch=1:nrow(dat), cex=cex)
abline(0,1)

# Ali-Submerged comparison
plot(dat[,"S"], dat[,"A"], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Submerged", ylab="Ali", main="IC50, Ali-Submerged")
abline(0,1)
mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))

# Submerged-Keratinocytes comparison
plot(dat[,"K"], dat[,"S"], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Submerged", main="IC50, Submerged-Keratinocytes")
abline(0,1)

# Ali-Fibroblasts comparison
plot(dat[,"F"], dat[,"A"], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts", ylab="Ali", main="IC50, Ali-Fibroblasts")
abline(0,1)

# Submerged-Fibroblasts comparison
# most alike
plot(dat[,"F"], dat[,"S"], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts", ylab="Submerged", main="IC50, Submerged-Fibroblasts")
abline(0,1)

# Ali-Keratinocytes comparison
plot(dat[,"K"], dat[,"A"], log="xy", pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Ali", main="IC50, Ali-Keratinocytes")
abline(0,1)
```


\newpage
For comparing 3D Ali and 3D Submerged, we fit a linear model for each drug. For other comparisons, which all involve more than 1 donor for some tissue models, we fit a linear mixed model with a donor random effect for each drug. 

2D vs 3D comes from comparing Keratinocytes and Fibroblasts with Submerged and Ali, using a mixed model adjusting for donor effects. Note that Submerged and Ali come from the same donor.

The overall p values come from comparing the four tissue models.

A table of the estimated ratios of IC50 (95\% CI) and adjusted p values are shown below. 

```{r, echo=F, message=F}

# turn Keratinocytes and Fibroblasts data into long format
dat=read.csv("fig5b.csv", row.names=1)

# i is 0 for Keratinocytes and 1 for Fibroblasts
dat.long=NULL
for (i in 0:1) {
  # reshape the first 3 columns into a long dataset
  idx=1:3+9*i
  dat.1 = myreshapelong(dat[,idx], cols.to.be.stacked=colnames(dat)[idx], label.cols.to.be.stacked="donor", new.col.name="id50")
  dat.1$drug = drugs
  dat.1$id=NULL
  dat.1$donor=3
   
  # repeat for the second set of 3 columns
  idx=4:6+9*i
  dat.2 = myreshapelong(dat[,idx], cols.to.be.stacked=colnames(dat)[idx], label.cols.to.be.stacked="donor", new.col.name="id50")
  dat.2$drug = drugs
  dat.2$id=NULL
  dat.2$donor=4
  
  # repeat for the third set of 3 columns
  idx=7:9+9*i
  dat.3 = myreshapelong(dat[,idx], cols.to.be.stacked=colnames(dat)[idx], label.cols.to.be.stacked="donor", new.col.name="id50")
  dat.3$drug = drugs
  dat.3$id=NULL
  dat.3$donor=5
  
  # rbind the three data frames 
  tmp = rbind(dat.1, dat.2, dat.3)
  tmp$tissue = ifelse(i==0, "Keratinocytes", "Fibroblasts")
  dat.long = rbind(dat.long, tmp)
}


# Turn ali and submerged data into long format
dat=read.csv("fig6a.csv", row.names=1)
# column 1 through 8 are submerged data, 9 through 13 are ali data
for (i in 1:8) {
  dat.long = rbind(dat.long, data.frame(donor="donor submerged/ali", id50=dat[,i], drug=rownames(dat), tissue="Submerged"))
}
for (i in 9:13) {
  dat.long = rbind(dat.long, data.frame(donor="donor submerged/ali", id50=dat[,i], drug=rownames(dat), tissue="Ali"))
}




# S-A comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d & tissue %in% c("Submerged", "Ali"))
  capture.output({fit = lm(log(id50) ~ tissue, data=tmp)}, type="message")
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=T)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_S_A=tab


# S-K comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d & tissue %in% c("Submerged", "Keratinocytes"))
  capture.output({fit = lmer(log(id50) ~ tissue + (1|donor), data=tmp)}, type="message")
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=T)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_S_K=tab


# A-F comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d & tissue %in% c("Ali", "Fibroblasts"))
  capture.output(suppressWarnings({fit = lmer(log(id50) ~ tissue + (1|donor), data=tmp)}), type="message")
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=T)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_A_F=tab


# K-F comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d & tissue %in% c("Fibroblasts", "Keratinocytes"))
  capture.output({fit = lmer(log(id50) ~ tissue + (1|donor), data=tmp)}, type="message")
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=T)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_K_F=tab


# 3D-2D comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d)
  tmp$Trt = ifelse(tmp$tissue %in% c("Fibroblasts", "Keratinocytes"), "2D", "3D")
  capture.output({fit = lmer(log(id50) ~ Trt + (1|donor), data=tmp)}, type="message")
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=T)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_3D_2D=tab



# All models comparison
pvals=sapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d)
  capture.output({fit = lmer(log(id50) ~ tissue + (1|donor), data=tmp)}, type="message")
  est=unlist(coef(fit)[[1]][1,-1])
  pchisq((est %*% solve(vcov(fit)[2:4,2:4]) %*% est)[1,1], 3, lower=F)
})
tab=p.adjust(pvals, method="holm")
names(tab)=rownames(dat)
tab_all=tab


# combine results
tab=cbind(tab_K_F, tab_S_A, tab_S_K, tab_A_F, tab_3D_2D, tab_all)
colnames(tab)=c("K-F", "adj.p", "S-A", "adj.p", "S-K", "adj.p", "F-A", "adj.p", "3D-2D", "adj.p", "Overall adj.p")
tab=tab[order(as.numeric(tab[,ncol(tab)])),]
print(tab)
mywrite.csv(tab, "fig6a_inferenc", row.names = T)

```



\newpage
## Fig 6b Four tissue models comparison - Max inhibition

Each point in the figure below represents the average of technical replicates for each donor. Thus, for Ali and Submerged, there is only one unique value for each drug.

```{r, echo=F, fig.width=9.6, fig.height=6}

dat=read.csv("fig5c.csv", row.names=1)
dat=round(dat, 0)
# take means across technical replicates
# dat.1 = sapply (1:6, function (i) (apply((dat[,1:3+(i-1)*3]), 1, mean, na.rm=T))) 
# take means across technical and donor replicates
dat.1 = sapply (1:2, function (i) (apply((dat[,1:9+(i-1)*9]), 1, mean, na.rm=T))) 

# add ali and submerged data
dat=read.csv("fig6b.csv", row.names=1)
# take means across technical replicates
submerged.mean=(apply((dat[,1:8]), 1, mean, na.rm=T))
ali.mean=(apply((dat[,9:13]), 1, mean, na.rm=T))

dat=cbind(dat.1, 
          submerged.mean, 
          ali.mean)

colnames(dat)=c("K", 
                "F",
                "S",
                "A"
                )
mywrite.csv(dat, "fig6bplot", row.names = T)


lim=range(dat, na.rm=T)
cex=1.5

par(mfcol=c(2,3))
par(mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

# Fibroblasts-Keratinocytes comparison
plot(dat[,"K"], dat[,"F"], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Fibroblasts", main="Max Inhibition, Fibroblasts-Keratinocytes")
abline(0,1)

# Ali-Submerged comparison
plot(dat[,"S"], dat[,"A"], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Submerged", ylab="Ali", main="Max Inhibition, Ali-Submerged")
abline(0,1)
mylegend(rownames(dat), x=1, col=1:nrow(dat), pch=1:nrow(dat))

# Submerged-Keratinocytes comparison
plot(dat[,"K"], dat[,"S"], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Submerged", main="Max Inhibition, Submerged-Keratinocytes")
abline(0,1)

# Ali-Fibroblasts comparison
plot(dat[,"F"], dat[,"A"], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts", ylab="Ali", main="Max Inhibition, Ali-Fibroblasts")
abline(0,1)

# Submerged-Fibroblasts comparison
# most alike
plot(dat[,"F"], dat[,"S"], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Fibroblasts", ylab="Submerged", main="Max Inhibition, Submerged-Fibroblasts")
abline(0,1)

# Ali-Keratinocytes comparison
plot(dat[,"K"], dat[,"A"], pch=1:nrow(dat), col=1:nrow(dat), cex=cex, xlim=lim, ylim=lim, xlab="Keratinocytes", ylab="Ali", main="Max Inhibition, Ali-Keratinocytes")
abline(0,1)
```

\newpage
For comparing 3D Ali and 3D Submerged, we fit a linear model for each drug. For other comparisons, which all involve more than 1 donor for some tissue models, we fit a linear mixed model with a donor random effect for each drug. A table of the estimated differences and adjusted p values are shown below.


```{r, echo=F, message=F}

# no log and exp for max inhibition
exp=F

# turn Keratinocytes and Fibroblasts data into long format
dat=read.csv("fig5c.csv", row.names=1)
dat=round(dat, 0)

# i is 0 for Keratinocytes and 1 for Fibroblasts
dat.long=NULL
for (i in 0:1) {
  # reshape the first 3 columns into a long dataset
  idx=1:3+9*i
  dat.1 = myreshapelong(dat[,idx], cols.to.be.stacked=colnames(dat)[idx], label.cols.to.be.stacked="donor", new.col.name="maxinh")
  dat.1$drug = drugs
  dat.1$id=NULL
  dat.1$donor=3
  
  # repeat for the second set of 3 columns
  idx=4:6+9*i
  dat.2 = myreshapelong(dat[,idx], cols.to.be.stacked=colnames(dat)[idx], label.cols.to.be.stacked="donor", new.col.name="maxinh")
  dat.2$drug = drugs
  dat.2$id=NULL
  dat.2$donor=4
  
  # repeat for the third set of 3 columns
  idx=7:9+9*i
  dat.3 = myreshapelong(dat[,idx], cols.to.be.stacked=colnames(dat)[idx], label.cols.to.be.stacked="donor", new.col.name="maxinh")
  dat.3$drug = drugs
  dat.3$id=NULL
  dat.3$donor=5
  
  # rbind the three data frames 
  tmp = rbind(dat.1, dat.2, dat.3)
  tmp$tissue = ifelse(i==0, "Keratinocytes", "Fibroblasts")
  dat.long = rbind(dat.long, tmp)
}


# Turn ali and submerged data into long format
dat=read.csv("fig6b.csv", row.names=1)

# column 1 through 8 are submerged data, 9 through 13 are ali data
for (i in 1:8) {
  dat.long = rbind(dat.long, data.frame(donor="donor submerged/ali", maxinh=dat[,i], drug=rownames(dat), tissue="Submerged"))
}
for (i in 9:13) {
  dat.long = rbind(dat.long, data.frame(donor="donor submerged/ali", maxinh=dat[,i], drug=rownames(dat), tissue="Ali"))
}


# S-A comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d & tissue %in% c("Submerged", "Ali"))
  fit = lm(maxinh ~ tissue, data=tmp)
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=exp)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_S_A=tab


# S-K comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d & tissue %in% c("Submerged", "Keratinocytes"))
  capture.output({fit = lmer(maxinh ~ tissue + (1|donor), data=tmp)}, type="message")
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=exp)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_S_K=tab


# A-F comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d & tissue %in% c("Ali", "Fibroblasts"))
  capture.output({fit = lmer(maxinh ~ tissue + (1|donor), data=tmp)}, type="message")
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=exp)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_A_F=tab


# K-F comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d & tissue %in% c("Fibroblasts", "Keratinocytes"))
  capture.output({fit = lmer(maxinh ~ tissue + (1|donor), data=tmp)}, type="message")
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=exp)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_K_F=tab


# 3D-2D comparison
fits=lapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d)
  tmp$Trt = ifelse(tmp$tissue %in% c("Fibroblasts", "Keratinocytes"), "2D", "3D")
  capture.output({fit = lmer(maxinh ~ Trt + (1|donor), data=tmp)}, type="message")
  fit
})
tab=data.frame(getFormattedSummary(fits, robust=F, type=3, exp=exp)[2,],
               getFormattedSummary(fits, robust=F, type=11, p.adj.method="holm", rows=2))
colnames(tab)=c("est (lb, ub)", "adjusted P value")
tab[[2]]=as.numeric(tab[[2]])
tab_3D_2D=tab



# All models comparison
pvals=sapply (drugs, function (d) {
  tmp=subset(dat.long, drug==d)
  capture.output({fit = lmer(maxinh ~ tissue + (1|donor), data=tmp)}, type="message")
  est=unlist(coef(fit)[[1]][1,-1])
  pchisq((est %*% solve(vcov(fit)[2:4,2:4]) %*% est)[1,1], 3, lower=F)
})
tab=p.adjust(pvals, method="holm")
names(tab)=rownames(dat)
tab_all=tab


# combine results
tab=cbind(tab_K_F, tab_S_A, tab_S_K, tab_A_F, tab_3D_2D, tab_all)
colnames(tab)=c("K-F", "adj.p", "S-A", "adj.p", "S-K", "adj.p", "F-A", "adj.p", "3D-2D", "adj.p", "Overall adj.p")
tab=tab[order(as.numeric(tab[,ncol(tab)])),]
print(tab)
mywrite.csv(tab, "fig6b_inferenc", row.names = T)
```

\newpage

# GFP trajectory

Goal: Generate a trendline for each cell type and/or cell type per donor, then compare the exponential phase slope or equation characteristics between keratinocytes, fibroblasts, and vero cells


## Plot GFP trajectory for each cell type

```{r, echo=F, fig.height=10}
dat=read.csv("GFP_Data_Figure4_raw.csv")
names(dat)[1]="time"

lim=range(dat[,-1])

par(mfrow=c(4,3), mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

# 6 donors for Keratinocytes
for (i in 1:6) {
  tmp=dat[,startsWith(colnames(dat), "K"%.%i)]
  mymatplot(dat$time, tmp, type="b", xlab="Time (hours)", ylab="GFP", main="Keratinocytes, Donor "%.%i, log="y", ylim=lim, pch=1, col=1:ncol(tmp), make.legend = F)
  abline(h=c(5e3, 18e3, 5e4), lty=c(2,1,2), col=c(1,"gold",1))
}

# 6 donors for Fibroblasts
for (i in 1:6) {
  tmp=dat[,startsWith(colnames(dat), "F"%.%i)]
  mymatplot(dat$time, tmp, type="b", xlab="Time (hours)", ylab="GFP", main="Fibroblasts, Donor "%.%i, log="y", ylim=lim, pch=1, col=1:ncol(tmp), make.legend = F)
  abline(h=c(5e3, 18e3, 5e4), lty=c(2,1,2), col=c(1,"gold",1))
}

```

```{r, echo=F, fig.height=10}
# Vero
par(mfrow=c(4,3), mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))
plot(V~time, data=dat, type="b", xlab="Time (hours)", ylab="GFP", main="Vero", log="y", ylim=lim)
lines(dat$time, dat[,"V.1"], col=2, type="b")
lines(dat$time, dat[,"V.2"], col=3, type="b")
lines(dat$time, dat[,"V.3"], col=4, type="b")
lines(dat$time, dat[,"V.4"], col=5, type="b")
abline(h=c(5e3, 18e3, 5e4), lty=c(2,1,2), col=c(1,"gold",1))

```

## Plot GFP percentage for each cell type

```{r, echo=F, fig.height=10}
dat=read.csv("GFP_Data_Figure4_pct.csv")
names(dat)[1]="time"

par(mfrow=c(4,3), mar=c(3.1, 3.1, 2.1, 1.1), mgp = c(2, 1, 0))

lim=c(0,100)

plot(Vero~time, data=dat, type="b", xlab="Time (hours)", ylab="GFP", main="Vero")
lines(dat$time, dat[,3], col=2, type="b")
lines(dat$time, dat[,4], col=3, type="b")
lines(dat$time, dat[,5], col=4, type="b")
lines(dat$time, dat[,6], col=5, type="b")

plot(Donor.1.Keratinocytes~time, data=dat, type="b", xlab="Time (hours)", ylab="GFP", main="Keratinocytes Donor 1", ylim=lim)
lines(dat$time, dat[,8], col=2, type="b")
lines(dat$time, dat[,9], col=3, type="b")
lines(dat$time, dat[,10], col=4, type="b")
lines(dat$time, dat[,11], col=5, type="b")

plot(Donor.2.Keratinocytes~time, data=dat, type="b", xlab="Time (hours)", ylab="GFP", main="Keratinocytes Donor 2", ylim=lim)
lines(dat$time, dat[,13], col=2, type="b")
lines(dat$time, dat[,14], col=3, type="b")
lines(dat$time, dat[,15], col=4, type="b")
lines(dat$time, dat[,16], col=5, type="b")

```
\newpage

## Estimate the exponential phase slope for each cell type and donor

Define the exponential phase as the time period when GFP crosses 18000. We use a simple algorithm which works under the assumption that the curve crosses this threshold only once, which is satisfied by the data. 

```{r, echo=F}

dat=read.csv("GFP_Data_Figure4_raw.csv")
names(dat)[1]="time"

# make sure dat$time is two hours apart
stopifnot(all(2==diff(dat$time)))

# For each column in dat[,-1], find the two values that sandwich 18000 and compute a slope
res=NULL
for (i in 2:ncol(dat)) {
  tmp=dat[,i]
  idx=which(tmp>18000)[1]
  if (length(idx)==0) next
  x=dat$time[idx:(idx-1)]
  y=tmp[idx:(idx-1)]
  slope=(log2(y[2])-log2(y[1]))/(x[2]-x[1])
  res=c(res, slope)
}
names(res)=names(dat)[-1]
cat("Slopes. V-V.4 are Vero replicates, K1-K1.4 are Keratinocytes donor 1 replicates, etc.\n")
print(res)

# putting results together
out=NULL

tmp = mean(res[startsWith(names(res), "V")])
names(tmp) = "V"
out=c(out, tmp)

tmp1 = sapply (1:6, function (i) mean(res[startsWith(names(res), "K"%.%i)]))
names(tmp1) = "K"%.%1:6
out=c(out, tmp1)

tmp2 = sapply (1:6, function (i) mean(res[startsWith(names(res), "F"%.%i)]))
names(tmp2) = "F"%.%1:6
out=c(out, tmp2)


out=cbinduneven(list(tmp, cbind(tmp1, tmp2)))
colnames(out)=c("Vero", "Keratinocytes", "Fibroblasts")
cat("Average slopes across biological replicates. There are 6 donors and one cell line (Vero).\n")
print(out)
mywrite.csv(out, "log2GFP_per_hr", row.names = T)
```

Paired Wilcoxon test between keratinocytes and fibroblasts
```{r, echo=F}
wilcox.test(tmp1, tmp2, paired=T)
```

Unpaired Wilcoxon test between keratinocytes and fibroblasts
```{r, echo=F}
wilcox.test(tmp1, tmp2, paired=F)
myboxplot(list(tmp1, tmp2), main="", xlab="", ylab="GFP slope", names=c("Keratinocytes", "Fibroblasts"), test="w")
```

Students' t test between keratinocytes and fibroblasts. Paired and then unpaired
```{r, echo=F}
t.test(tmp1, tmp2, paired=T)
t.test(tmp1, tmp2, paired=F)
```



Unpaired Wilcoxon test and Student's t test (assuming equal variance) between vero and keratinocytes
```{r}
wilcox.test(tmp, tmp1, paired=F)
myboxplot(list(tmp, tmp1), main="", xlab="", ylab="GFP slope", names=c("Vero", "Keratinocytes"), test="w")
# t test using two different degrees of freedom
pt((mean(tmp1)-mean(tmp))/(sqrt(var(tmp1)/6)*7), lower.tail=F, df=5)
pt((mean(tmp1)-mean(tmp))/(sqrt(var(tmp1)/6)*7), lower.tail=F, df=2)
```

Unpaired Wilcoxon test and Student's t test (assuming equal variance) between vero and fibroblasts
```{r}
wilcox.test(tmp, tmp2, paired=F)
myboxplot(list(tmp, tmp2), main="", xlab="", ylab="GFP slope", names=c("Vero", "Fibroblasts"), test="w")
# t test
pt((mean(tmp2)-mean(tmp))/(sqrt(var(tmp2)/6)*2), lower.tail=F, df=5)
pnorm((mean(tmp2)-mean(tmp))/(sqrt(var(tmp2)/6)*2), lower.tail=F)
```